A Theory for Migration of Information Systems

### 2022-09-04 2022 Migration/articleMigration.tex
WIP

\section{Basics}
\label{sct:Basics}
	To define migration, we must introduce the notion of dataset because migration aims to preserve data whilst changing the software that manipulates data.
	A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.
	To abstract away from a particular database implementation,
	we represent data elements as triples.
	This makes our theory valid for any kind of database that triples can represent,
	such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases

	Before defining datasets, we must first define the constituent notions of atom, concept, relation, rule, and role.
\subsection{Atoms}
	Atoms are values that have no internal structure, meant to represent data elements in a database.
	From a business perspective, atoms are used to represent concrete items of the world,
	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
	The set of atoms is called $\atoms$.

@@

\subsection{Triples}
	A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean that the ``thing'' that \atom{Peter} refers to
	has \atom{074238991} as a telephone number.
	Triples are used to relate two atoms and a relation.
	In the formal world, we leave it entirely up to a user to attach meaning to a triple.
	This ``meaning from practice'' has no consequences in the formal world.
	The set of triples is called $\triples$.

@@

	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concept symbols $A$ and $B$.
	It contains all pairs that can be made of elements of $A$ and $B$:
\[\pop{\fullt{A}{B}} = \{ \pair{a}{b}| a\ \inst\ A, \}\]
	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
	The set of relations is called $\rels$.

@@

\subsection{Constraints}
	Each constraint $u\in\rules$ has a signature $\signat{u}$, which is a pair of concepts.
	Each constraint $u$ also has a function $\id{viol}_u:\powerset{\atoms\times\rels\times\atoms}\ \rightarrow\ \powerset{\atoms\times\atoms}$.
	We call the set $\viol{\triples}{u}$ the \define{violation set} of constraint $u$ in population $\triples$.
	We call every element $\pair{a}{b}\in\viol{\triples}{u}$ a \define{violation} of constraint $u$ in $\triples$.
	Note that the signature of $u$ does not depend on the population, whereas the violation set of $u$ may be different for every $\triples$.
	The set of constraints is called $\rules$.

\subsection{Datasets}
	We can now define the notion of dataset.
	Let	$\concepts$ be a set of concept symbols,
	$\rels$ be a set of relations,
	$\atoms$ be a set of atoms,
	let $\inst$ be a relation $\atoms\times\concepts$ that represents the instance relation between concepts and atoms,
	% let $\roles$ be a set of roles,
	and $\rules$ be a set of constraints.
\begin{definition}[dataset]
	\label{def:dataset}
	\item A dataset is a tuple $\la\concepts,\isa,\rels,\atoms,\inst,\triples,\rules\ra$ with
	\[\begin{array}[3]{rcl}
		% \id{maintain}&:&\roles\times\rules\\
		\isa\text{\ is a partial order}\\
		\forall\triple{a}{r}{b}\in\triples&:&a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}\\
		\forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
		\forall u\in\rules&:&\viol{\triples}{u}=\emptyset
	\end{array}\]
\end{definition}

\subsection{Motivations}
	We have already motivated the use of triples in a dataset:
	It is a generic way of representing structured data, which maps to most (if not all) existing database technologies.
	Changes in the data means changing the set of triples, adding or removing triples from the set.
	That is not migration; that is regular operational behaviour.
	If the software changes, this means that the concepts, relations, and constraints change as well.
	However, a dataset can contain atoms that do not partake in any triple.
	That is why we need a set of atoms, $\atoms$ too.

	Atoms serve as data elements. Changes in the dataset are reflected in $\atoms$.
	Concepts serve to classify atoms in a partial order (the classification) so we can implement a typing regime on the dataset.
	This type system ensures that the dataset can be interpreted as a meaningful database.
	As a consequence, the relations $\isa$ and $\inst$

	The choice of elements in the tuple that represents a dataset is determined by the dynamics of a dataset.
	Anything that can change needs to be in that tuple.
	It is not only the data that changes. Rules can change, relations and concepts can change, the taxonomy can change,
	rules can change.
	In migrations we want to cope with any conceivable change,
	ensuring that the dataset satisfies its defining properties (from definition~\ref{def:dataset}).
	The migration must also ensure that the population remains intact as much as possible.

>>>>

\section{Basics}
\label{sct:Basics}
	To define migration, we must introduce the notion of dataset
	because migration aims to preserve data while changing the software that manipulates data.
	A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.
	To abstract away from a particular database implementation,
	we represent data elements as triples.
	This makes our theory valid for any kind of database that triples can represent,
	such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases

	Before defining datasets, we must first define the constituent notions of atom, concept, relation, rule, and role.
\subsection{Atoms}
	Atoms serve as data elements.
	They are values without internal structure of interest, meant to represent data elements in a database.
	From a business perspective, atoms are used to represent concrete items of the world,
	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
	The set of atoms is called $\atoms$.

@@

\subsection{Triples}
	Triples are used to represent data.
	Each triple relates two atoms to a relation.
	A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean that the ``thing'' that \atom{Peter} refers to
	has \atom{074238991} as a telephone number.
	In the formal world, we leave it entirely up to a user to attach meaning to a triple.
	This ``meaning from practice'' has no consequences in the formal world.
	The set of triples is called $\triples$.

@@

	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concept symbols $A$ and $B$.
	It contains all pairs that can be made of elements of $A$ and $B$:
\[\pop{\fullt{A}{B}} = \{ \pair{a}{b}| a\ \inst\ A,\ b\ \inst\ B \}\]
	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
	The set of relations is called $\rels$.

@@

\subsection{Constraints}
	Constraints are the invariants of a dataset: restrictions that are kept true at all times.
	Their purpose is to ensure the semantics of a dataset, i.e. the preservation of meaning as data changes.
	Each constraint $u\in\rules$ has a signature $\signat{u}$, which is a pair of concepts.
	Each constraint $u$ also has a function $\id{viol}_u:\powerset{\atoms\times\rels\times\atoms}\ \rightarrow\ \powerset{\atoms\times\atoms}$.
	We call the set $\viol{\triples}{u}$ the \define{violation set} of constraint $u$ in population $\triples$.
	We call every element $\pair{a}{b}\in\viol{\triples}{u}$ a \define{violation} of constraint $u$ in $\triples$.
	Note that the signature of $u$ does not depend on the population, whereas the violation set of $u$ may be different for every $\triples$.
	The set of constraints is called $\rules$.

\subsection{Datasets}
	We can now define the notion of dataset.
	Let	$\concepts$ be a set of concept symbols,
	let $\isa$ be a partial order on $\concepts$,
	let $\rels$ be a set of relations,
	let $\atoms$ be a set of atoms,
	let $\inst$ be a relation $\atoms\times\concepts$ that represents the instance relation between concepts and atoms,
	% let $\roles$ be a set of roles,
	and let $\rules$ be a set of constraints.
\begin{definition}[dataset]
	\label{def:dataset}
	\item A dataset is a tuple $\la\concepts,\isa,\rels,\atoms,\inst,\triples,\rules\ra$ with
	\begin{eqnarray}
		\forall r\in\rels&:&\src{r}\in\concepts\ \wedge\ \tgt{r}\in\concepts\\
		\forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
		% \id{maintain}&:&\roles\times\rules\\
		\forall A,B\in\concepts, a\in\atoms&:&A\isa B\wedge a\ \inst\ A\ \Rightarrow\ a\ \inst\ B\\
		\forall\triple{a}{r}{b}\in\triples&:&r\in\rels\ \wedge\ a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}\\
		\forall u\in\rules&:&\viol{\triples}{u}=\emptyset\label{violation free}
	\end{eqnarray}
\end{definition}

\subsection{Motivation}
% The purpose of this section is to explain why a dataset is structured the way it is.
	The challenge of migration is to preserve meaning rather than just copy data from one place to another.
	The bearer of meaning to users of a dataset is the data.
	That is represented by the population, i.e. the sets of triples and atoms, $\triples$ and $\atoms$.
	How a user interprets the data is up to the user and may therefore have no consequence to the dataset.
	So we attach no intrinsic meaning to a triple and take $\triples$ and $\atoms$ for granted.
	In a migration we seek to preserve that data where we can.
	The other elements in a dataset, $\isa$, $\concepts$, $\rels$, and $\rules$ do have consequences, however.
	Definition~\ref{def:dataset} implies that changes to data may cause violations of rules,
	so measures must be taken to get rid of those violations (eqn.~\ref{violation free}).
	Changes to $\isa$, $\concepts$, $\rels$ change the data structure in a dataset.
	Changes to $\rules$ change the invariants of the dataset.
	This paper studies those changes and seeks to preserve the rules in definition~\ref{def:dataset} in such a way that large portions of migration can be automated.
